Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan

### 2025-03-09 analysis/paper/paper.qmd
draft editing

<!-- NE Taiwan as a case study -->
While many studies focus on prominent colonies, places on the periphery of colonial bases were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017]. Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion. Yilan, located in northeastern Taiwan, is an example of a remote region where European control was limited and indirect for most of the time but was intensively involved in regional trade activities with foreign groups. In the mid-17th century, the VOC (Dutch East India Company) monopolized the deerskin trade in Taiwan for export on a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001; @Cheng2017]. Despite many discussions on the operations and policies of the deerskin trade through Dutch and Chinese written records [@Lin2014deer], direct investigations on the colonial experiences of local indigenous people through archaeological evidence are rare. The interaction between Indigenous people and their environments in the colonial setting of economic demands requires more Indigenous perspectives from archaeological studies.

<!-- research aim and questions -->
In this paper, I will investigate the Indigenous subsistence in a colonial context through the case study of the Kiwulan site (AD 1350-1950), one of the largest Iron Age settlements in northeastern Taiwan [@Chen2007]. Kiwulan encountered European colonial powers, including the Spanish and Dutch in the 17th century, followed by a large influx of Han Chinese migration in the 19th century, making it a suitable example to explore the reactions of Indigenous groups to the foreign presence over time. My research questions are: What was the resource use of Indigenous people before and after those foreign presences, especially during the Dutch period with large-scale trade? Can we identify the evidence of getting involved in the deer trade from faunal remains? If so, were there subsistence strategies in response to the impacts of economic demands by the Europeans and Chinese? To address these questions, I will analyze the faunal remains with a focus on vertebrate species from Kiwulan because zooarchaeology is one useful approach to studying local consumption and exchange systems [@Defrance2009; @Crabtree1990]. Through zooarchaeological evidence, this study aims to stress the active role of Indigenous people at the periphery of colonial bases in regional trade networks.

# The European colonization period and regional trade network in Northern Taiwan
<!-- Spanish colonization and the power relations in particular spatial and temporal contexts-->
In 1626, the Spanish (Spanish East Indies) established the first colony of San Salvador at Heping Dao, and later Santo Domingo in Tamsui in northern Taiwan (@fig-KWL-map), to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007]. Although the Spanish intended to build a colonial city, San Salvador remains its early development during their colonization, primarily consisting of a fortress and a church [@Borao2009, pp. 103-134; @Berrocal2018]. The neighboring area of the fortress was inhabited by two Indigenous groups, the Taparris and the Quimaurris, who had lived at the fort location before being expelled by the Spanish [@Borao2001, SIT:165, p.81]. The relations between the Spanish and Indigenous people might not involve extensive colonialism or intercultural domination, with limited territorial control only to the fort. Also, the Spanish relied on the exchange of local goods and food surpluses with Indigenous people, according to written records [@Borao2001, SIT:170-171]. This corresponds with a higher proportion of Indigenous and Chinese items with little European materials found in San Salvador [@Berrocal2018], indicating cultural entanglement.

<!-- the Dutch domination, how and why colonial powers were able to exert control -->
In 1642, the Dutch defeated the Spanish and took over San Salvador and Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662. Compared to the Spanish, the Dutch was more strategic in consolidating their power by establishing political and economic institutions, forming alliances with Indigenous communities, and conquering resistant communities [@Kang2016; @Andrade2007]. For examples, the colonial institutions to Indigenous people include taxation, tribute system, village-leasehold system, census investigations, annual gatherings (Dutch: Landdag), and Indigenous representative from each aliened village [@Kang2016; @Andrade1997; @Andrade2007]. Among those institutions, the tribute system operated from 1642 to 1647 and the village-leasehold system introduced in 1644 had great impacts on Indigenous economy and the environment they lived in. Based on the tribute system, each household in allied communities was required to pay rice or deer products annually, especially deerskin. The amount of deerskin collected from Indigenous communities in northern Taiwan once reached 7,000 pieces in 1646 [@Andrade2007]. Moreover, to increase deerskin exports to Japan, the Dutch implemented a hunting license system (1636-1645) that gave Han Chinese permission to hunt deer on Indigenous lands during certain periods [@Heyns2001]. From 1633 to 1661, the average number of deerskins exported from Taiwan was about 67,000 per year [@Koo2011; @Cheng2017]. Those colonial activities might affect the economy of indigenous people since deer was an important resource in prehistoric Taiwan [@Chen2000animal].

<!--variation of colonial powers across regions and indirect colonization at Kavalan-->
Although the Dutch power expanded across nearly whole Taiwan island in the mid-17th century, the extent of their colonial dominance differed greatly from region to region. Colonial power over regions far from the European fortress were usually difficult and unsuccessful. Yilan in northeastern Taiwan is an example of minimal European dominance or indirect colonization for most of the time [@Kang2012]. As an plain located around 70 kilometers away from the European colonies with mountains in between, Yilan was hardly accessed by land but mostly from sea. The main Indigenous group in Yilan was the Kavalan. Given its strategic location on the route to Torrobouan, a place believed having gold mines, the Dutch was eager to integrate Yilan under their control [@DZI2000; @Kang2016, p47]. In 1644, the Dutch defeated several Kavalan communities and incorporated them into their colonial territory [@Kang2016]. Instead of engaging directly with the Kavalan people, the Dutch relied mostly on the assistance of the mediate, the Basay people active in northern Taiwan [@Kang2003], to coordinate and collect tributes from Yilan. This indirect approach allowed the maintenance of autonomy of the Kavalan people, and was even able to negotiate tributes requested by the VOC [@Kang2016].

<!-- the pre-existing trade network in norththern Taiwan-->
Before the arrival of the Europeans in northern Taiwan, an inter-insular trade network had been established since the 14th century, involving Chinese merchants and indigenous groups in Tamsui and Keelung [@Chen2005]. Chinese traders transported goods such as textiles, porcelains, beads, and metal items to exchange for raw materials from Taiwan, such as gold, sulfur, and animal hides and meats from the Basay people [@Chen2005; @Andrade2007]. This regional trade network was further connected to maritime trade between Ming China and the Ryukyu Kingdom during the 15th and 16th centuries and later to European long-distance trade in the 17th century [@Chen2005; @Von2019; @Wang2007]. Yilan was included in this long-established trade network, where the Kavalan people exchanged products such as rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003]. This is evidenced by the large amount of imported foreign goods found across archaeological sites in Yilan, indicating frequent trade activities, especially around the 17th century [@TaiwanHistory2005]. This trade network was then intensified during the European presence, and the participation of Kavalan in this network was driven by the pursuit of foreign goods, and also likely colonial pressure from European powers. This unique historical and spatial context makes Yilan an ideal location to study Indigenous reactions to foreign influences and the associated trade networks from a long-term perspective in a pericolonial context [@Acabado2017; @Kang2016, pp. 31-78; @Trabert2017].

# Archaeological Context at Kiwulan
<!--introduction of Kiwulan and the population-->
Kiwulan, located in the northern part of Yilan (@fig-KWL-map), is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. A large scale salvage excavation was conducted at Kiwulan that exposed a total of 262 excavation units (4 * 4 m), revealing abundant artifacts and numerous features, including middens, burials, ditches, and post-holes with wooden posts in situ. The excavtion suggests that Kiwulan was a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004; @Chiu2004]. The main inhabitants of Kiwulan were the Kavalan people based on oral history and the similarities between Kavalan material culture recorded in historical records and archaeological artifacts [@Chen2004]. Kavalan people is the main Indigenous group in Yilan during the late Iron Age and the early modern period. According to a Dutch census conduced in the mid-17th century, there were about 40 Kavalan settlements with approximately 10,000 people in Yilan. Kiwulan, written as "Kibannoran" in Dutch records, was the largest settlement, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which eventrually led to the relocation of Kavalan people to Hualien in eastern Taiwan [@Zhan1995].

<!--archaeological materials-->
The most common artifact found at Kiwulan was locally made pottery with geometric patterns as surface treatments, which was also found at many simultaneous sites in Yilan [@TaiwanHistory2005]. In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts and ornaments, were largely found in both the residential area and burials. Due to no evidence of metallurgy and local production of beads, these items are believed to have been acquired through exchange, primarily with the Basay people active in northern Taiwan. This indicates that Kiwulan has continuously participated in regional trade networks closely associated with northern Taiwan since the 14th century and influenced by the long-distance trade during the 17th century [@Wang2007]. Moreover, the use of foreign goods appears to have cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021; @Cheng2008].

<!--archaeological remains-->
Abundant animal bones, shellfish, and plant remains were uncovered at Kiwulan, reflecting the diversity of natural resources in the plain and an environment close to estuary. The animal resource includes deer, suids, cattle, birds, fish, shellfish, and plants mainly consisting of rice, peach, pandanus, and cucurbits [@Chen2004]. Rice farming seems to be common but likely limited to a household scale rather than intensive agriculture according to historical records [@Chen1963]. Shellfish consist of bivalves such as clams and oysters and univalves from rivers and brackish water environments. Based on those archaeological remains and environments, Kiwulan residents should have practiced a mixed subsistence that includes rice farming, hunting, and fishing for wild resources. Artifacts associated with those subsistence practices were also found at Kiwulan, such as iron arrowheads for hunting, iron sickles for crop harvesting, and wooden paddles for water-related activities. However, the resource use across different time periods still need further investigations to better understands the subsistence adaptation relative to foreign presence and environmental changes.

@@

To construct a reliable fine-grained chronological sequence, archaeological deposits were divided into six cultural layers (CL6-CL1) spanning from the 14th century to the 19th century based on excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators. The diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks relating to the 19th century [@Wang2011; @Hsieh2009]. Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above could still offer a relatively robust framework for comparing archaeological records before and after the foreign influence. Among these six temporal layers, the fourth cultural layer (CL4) can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second cultural layer (CL2) corresponds to the 19th century, associated with the large influx of Chinese migrants in Yilan.

@@

# Discussion
<!--results and chronology -->
The sampled faunal assemblage from the major residential area at Kiwulan suggests a continuation of sika deer consumption from CL6 to CL2, and then shifted to bovine in CL1. Despite deer was a substantial part of subsistence at Kiwulan, variations in their quantity, ubiquity, butchering patterns, and proportions relative to other animals show changes that might relative to foreign impacts. Based on radiocarbon dates and diagnostic artifacts indicating specific time [@Wang2020ornament; @Chen2007], the cultural layers were further grouped into temporal sequences in relation to foreign presences for further discussion. The temporal sequences include the pre-European period (CL6, CL5), the European period (CL4), the post-European period (CL3), and the arrival of Han-Chinese immigrants (CL2, CL1). It should be noted that the European period does not mean a solid European colonial control over Kiwulan, but a broad impact from European occupation in Northern Taiwan and associated colonial economic activities on indigenous people in Yilan, such as getting involved in the long-distance trade network, tribute paying, and suffered from occasional European attacks. The small sample size after separating them into different periods might limits a comprehensive interpretation. However, comparing the results in a continuous sequences still provide evident patterns reflecting changes in animal use.

<!--summary of overall observations across time periods -->
The sampled area at Kiwulan shows a large quantity of faunal remains in the European period, followed by the late pre-European period. Interestingly, deer bones from the late pre-European period presented the same ubiquity as in the European period but in only half the quantity. This might indicate consistency in the  regularity of deer consumption at the settlement scale before and after European presence, but deer utilization was intensified in the European period. During the European period, in addition to the predominance of deer, an increased relative proportion and ubiquity of suid suggests a more regular use of suid than in other periods. The dramatic decrease in deer bones after the European period might be related to both the decline of the Kiwulan population and a decrease in the wild deer population. Sika deer populations in western Taiwan have significantly shrunk since the 18th century because of habitat loss and overhunting [@Chen2000animal]. An increase in muntjac use may suggest they served as a substitute, replacing the role of suids in earlier periods. Despite this, deer remained a predominant resource for local consumption at Kiwulan during the post-European and early Chinese period. Bovine appeared in the pre-European period with just a few evidence (yellow cattle), but became the only animal (water buffalo) used at Kiwulan in the late Chinese period. Since bovine was not domestic aniamls that were beleived introduced from mainland China, this suggests a Chinese influence and the relocation of the Kavalan people to Hualien due to population pressures from Chinese immigration [@Zhan1995].

<!--skeletal part frequencies and butchery patterns of deer-->
Although deer was the primary resource at Kiwulan, differences in skeletal completeness across periods suggests changes in butchery locations and cooking practices. There were more complete carcasses in the European and late pre-European period compared to the post-European and Chinese periods. Moreover, the substantial rise in fragmented head parts (cranium and mandible) during the European period indicates greater use of soft tissues, such as brains and tongues, or the transport of more intact deer to the settlement for processing. This aligns with the historical account in 1603, Record of Formosa (Dongfan ji) [@Chen1603], which describes deer tongue consumption and group hunting for deer in western Taiwan. Changes in cooking practices are also evident. The decline in burned bones over time with a lowest percentage in the European period suggests a shift from roasting to other cooking methods, such as boiling. The presence of small chopped bones that could fit in local ceramics at Kiwulan indicates boiling as a possible practice since the European period. This could be further confirmed through organic residue analysis on pottery in the future. Other meat preparation such as drying (or salting and smoking) cannot be ruled out [e.g., @Spyrou2019]. Although deer vertebrae and ribs, portions potentially used for drying [@Friesen2001; @Binford1978], were in low frequencies at Kiwulan, dry meat consumption or exchange by the Kavalan people was reported in historical records [@Chen1963]. The shift in cooking practices might result from increased need for deer, which could have led people to adopt boiling due to its additional benefits, such as maximizing residual meat use, facilitating grease rendering, and improving quality of lean meat [@Lupo1997], and drying for meat preserving and transporting.

A consistent pattern in deer part frequencies was also identified throughout all periods. There were high frequencies of upper limb bones but low frequencies of vertebrae and ribs, suggesting selectively butchered and preferences of meaty portions. Moreover, among those meaty portions, front quarters (humerus and radius) are always more abundant than hindquarters (femur and tibia). One possible explanation is that femurs and tibiae were heavily fragmented for marrow extraction due to their high yields [@Madrigal2002; @Lupo1998], making them less identifiable in the assemblage. Marrow extraction can be further supported by at least 25% of long bones showing fractures and a moderate degree of fragmentation in general. Another likelihood is that some hindquarters might be used for trading, as Dutch records mention deer legs and meats as commodities in addition to hides [@Blusse2006, pp.220-224], but without specifying which legs. Deer legs imply bone-meat packages that would result in a low frequencies of bones of hindquarters or front quarters. Considering the overall high proportion of meaty portions at Kiwulan, deer were likely processed not only for local consumption but also for trade. The lower frequencies of hindquarters than front quarters might be attributed to both marrow extraction and export of deer products, rather than a sole contribution.

<!--deer hides-->
Deerskin trade had prevailed since the 14th century in East Asia, and monopolized by the Europeans in the 17th century, when large hides were exported to Japan, especially from western Taiwan [@Cheng2017; @Hollmann1991; @Wang1981]. The cut marks identified on deer lower limbs at Kiwulan confirm the removal of deer skins and likely being traded with external traders. Abundant foreign goods at Kiwulan, including ornaments, porcelains, and metal items, supports frequent exchanges between local residents and outsiders [@Wang2020ornament; @Chen2007]. This suggests that not only western Taiwan but also northeastern Taiwan was intensively involved and impacted by the long-distance trade. Although lower limbs with skinning cuts did not show a substantial increase in the European period, the disproportionate increase in head parts might hint at a likelihood of brain tanning when preparing hides in addition to tissue consumption. Brain tanning is a traditional way for softening and preserving hides in many Indigenous cultures, such as North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016]. The present-day Bunun group in the central highlands of Taiwan has revived hide tanning based on oral history, primarily using peanuts or pig lard as tannnig materials, with a few examples using animal brains [@Li1999, pp. 48; @Xu1994]. Despite the materials for tanning used by the Kavalan people being unknown due to the loss of traditions, brain tanning is a practicable method given the availability of deer heads from hunting.

>>>>

<!-- NE Taiwan as a case study -->
While many studies focus on major colonial centers, peripheral regions also experienced varying degrees of European influence through colonial powers and associated trade networks, namely pericolonialism [@Acabado2017; @Trabert2017]. Exploring these areas is important to for understanding the dirverse impacts of colonial expansion. Yilan, in northeastern Taiwan, is an example of a remote region where European control was limited and indirect but Indigenous communities were deeply involved in regional trade with foreign groups. In the mid-17th century, the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) monopolized the deerskin trade in Taiwan and exported deerskin on a large scale to Japan [@Heyns2001; @Cheng2017]. While historical records by the Dutch and Chinese provide details about the operations and policies of the deerskin trade [@Lin2014deer], archaeological investigations into Indigenous reactions to these economic activities remain rare. To fully understand Indigenous responses to colonial economic demands, archaeological perspectives on their interactions with the environment are essential.

<!-- research aim and questions -->
In this paper, I investigate Indigenous subsistence in a pericolonial context through the case study of the Kiwulan site (AD 1350-1850), one of the largest Iron Age settlements in northeastern Taiwan [@Chen2007]. Kiwulan experienced European colonial powers, including the Spanish and Dutch in the 17th century, followed by a large influx of Han Chinese migrats in the 19th century, making it an ideal case to explore Indigenous responses to foreign presence over time. Key research questions include: How did Indigenous resource use change before and after these foreign encounters, particularly during the Dutch period of large-scale trade? Can we identify evidence of Indigenous participation in the deer trade from faunal remains? If so, any subsistence strategies emerged in response to economic demands from Europeans and Chinese traders? To address these questions, I analyze faunal remains with a focus on vertebrate species from Kiwulan, as zooarchaeology is an useful approach to studying local consumption and exchange systems [@Defrance2009; @Crabtree1990]. Through this approach, this research highlights the active role of Indigenous communities at the periphery of colonial bases within regional trade networks.

# The European colonization period and regional trade network in Northern Taiwan
<!-- Spanish colonization and the power relations in particular spatial and temporal contexts-->
In 1626, the Spanish (Spanish East Indies) established their first colony in northern Taiwan, San Salvador at Heping Dao, follwed by Santo Domingo in Tamsui (@fig-KWL-map), to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market. The VOC had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007]. Although the Spanish intended to build a colonial city, San Salvador remained in its early stages of development throughout their colonization, consisting primarily of a fortress and a church [@Borao2009, pp. 103-134; @Berrocal2018]. The area surrounding the fortress was inhabited by Indigenous groups, the Taparris and the Quimaurris, who had lived at the site before being expelled by the Spanish [@Borao2001, SIT:165, p.81]. The relations between the Spanish and Indigenous people may not have involved extensive colonial control or intercultural domination, as Spanish territorial control appears to have been limited to the fort. Instead, the Spanish relied on exchange of local goods and food surpluses with Indigenous communities based on historical documents [@Borao2001, SIT:170-171]. This is consistent with the archaeological evidence from San Salvador, which shows a higher proportion of Indigenous and Chinese artifacts, with relatively few European materials [@Berrocal2018], indicating cultural entanglement.

<!-- the Dutch domination, how and why colonial powers were able to exert control -->
In 1642, the Dutch defeated the Spanish and took control of San Salvador and Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662. Compared to the Spanish, the Dutch was more strategic in consolidating their power by establishing political and economic policies, forming alliances with Indigenous communities, and conquering resistant ones [@Kang2016; @Andrade2007]. Examples of colonial institutions imposed on Indigenous people include taxation, a tribute system, a village-leasehold system, census investigations, annual gatherings (Dutch: Landdag), and Indigenous representative selection [@Kang2016; @Andrade1997; @Andrade2007]. Among those colonial policies, the tribute system (1642-1647) and the village-leasehold system introduced in 1644 had profound impacts on Indigenous economies and the environment. Under the tribute system, each household in allied communities was required to pay deer products or rice to the Dutch annually, with deerskins being a primary item. In 1646, the amount of deerskin collected from Indigenous communities in northern Taiwan was 7,000 pieces [@Andrade2007]. Moreover, to increase deerskin yields, the Dutch implemented a hunting license system (1636-1645), which permitted Han Chinese to hunt deer on Indigenous lands during specific periods [@Heyns2001]. From 1633 to 1661, the average annual export of deerskins from Taiwan reached 67,000 [@Koo2011; @Cheng2017]. These colonial activities likely impacted the economy of Indigenous peoples, because deer was an important resource in prehistoric Taiwan [@Chen2000animal].

<!--variation of colonial powers across regions and indirect colonization at Kavalan-->
Although Dutch power expanded across nearly the entire Taiwan in the mid-17th century, the extent of their their colonial dominance varied greatly by region. Colonial control over areas far from the Dutch fortresses was often difficult and unsuccessful. Yilan in northeastern Taiwan is an example of weak colonial control or indirect colonization for most of the time [@Kang2012]. Yilan is located around 70 kilometers from European colonies with mountains in between, which made it difficult to access by land and was primarily reached by sea. In Yilan, Kavalan people was the main Indigenous group since the Iron Age. Due to the strategic position of Yilan on the route to Torrobouan, a place believed to have gold mines, the Dutch were eager to bring Yilan under their control [@DZI2000; @Kang2016, p47]. In 1644, the Dutch defeated several Kavalan communities and incorporated them into their colonial territory [@Kang2016]. However, instead of directly engaging with the Kavalan people, the Dutch relied on the assistance of intermediaries, the Basay people, who were active in Northern Taiwan, to collect tributes [@Kang2003]. This approach of indirect control allowed the Kavalan to maintain a degree of autonomy and even negotiate tributes [@Kang2016].

<!-- the pre-existing trade network in norththern Taiwan-->
Before the arrival of Europeans, an inter-insular trade network had been established since the 15th century in northern Taiwan, involving Chinese merchants and Indigenous groups in Tamsui and Keelung [@Chen2005]. Chinese traders transported goods such as textiles, porcelains, beads, and metal items in exchange for raw materials from Taiwan, including gold, sulfur, rattan, and animal hides and meats [@Chen2005; @Andrade2007]. This regional trade network was further connected to maritime trade between Ming China and the Ryukyu Kingdom around 15-16th centuries and later integrated into European long-distance trade in the 17th century [@Chen2005; @Von2019; @Wang2007]. Yilan was part of this pre-existing trade network, where local Kavalan people exchanged rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003]. This trade network intensified during the period of European presence, which is consistent with the large quantities of imported foreign goods found at archaeological sites across Yilan, indicating frequent trade activity in the 17th century [@TaiwanHistory2005]. This unique historical and spatial context makes Yilan an ideal location for exploring Indigenous responses to foreign presence and the dynamics of trade networks in a pericolonial context [@Acabado2017; @Kang2016, pp. 31-78; @Trabert2017].

# Archaeological Context at Kiwulan
<!--introduction of Kiwulan and the population-->
Kiwulan, located in northern Yilan (@fig-KWL-map), is a multi-component site, with its upper component spanning from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. A large-scale salvage excavation at Kiwulan uncovered a total of 262 excavation units (4 x 4 m), revealing abundant artifacts and numerous features, including middens, burials, ditches, and post-holes with wooden posts in situ. The excavation suggests that Kiwulan was a well-developed settlement with longhouses (approximately 4 x 8 m) arranged side by side, and burials located to the north of the residential area [@Chen2004]. Based on oral history and the similarities between Kavalan material culture recorded in historical documents and the archaeological assemblage, Kiwulan is identified as a Kavalan settlement [@Chen2004]. The Kavalan people were the main Indigenous group in Yilan during the late Iron Age and early modern period. According to a Dutch census from the mid-17th century, there were approximately 40 Kavalan settlements in Yilan, with a total population of about 10,000 people [@Nakamura1938]. Among these Indigenous settlements, Kiwulan ("Kibannoran" in Dutch sources) was the largest one with 160 households and 711 inhabitants. In the early 19th century, increasing Han Chinese migration gradually led to the relocation of the Kavalan people to Hualien in eastern Taiwan [@Zhan1995]. This is reflected in the dramatic decline of Kiwulan’s population to fewer than 100.

<!--archaeological materials-->
The most common artifact found at Kiwulan was locally made pottery with geometric surface patterns, a typical type of ceramic in Yilan [@TaiwanHistory2005]. In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts and ornaments, were largely found in both residential area and burials. Since there is no evidence of local metallurgy or bead production, these items were likely obtained through exchange. This suggests that Kiwulan maintained continuous participation in regional trade networks linked to northern Taiwan, which were further influenced by long-distance trade in the 17th century [@Wang2007]. Moreover, foreign goods such as beads were considered prestige items or used for ritual purposes, and previous studies suggest that their introduction may have led to increasing social differentiation [@Wang2020ornament; @Wang2021; @Cheng2008].

<!--archaeological remains-->
Abundant animal bones, shellfish, and plant remains were uncovered at Kiwulan, reflecting the diversity of natural resources in the plain and an environment close to estuary. The faunal remains include deer, suids, cattle, birds, fish, and shellfish, while plant remains primarily consist of rice, peach, pandanus, and cucurbits [@Chen2004]. Rice farming appears to have been practiced, but likly limited a household scale rather than as intensive agriculture, according to archaeological evidence and historical records [@Chen1963]. The shellfish remains include bivalves such as clams and oysters, and univalves from freshwater and brackish water environments. Based on the archaeological remains and local environment, it is likely that Kiwulan residents practiced a mixed subsistence strategy, including rice farming, hunting, and fishing. Artifacts associated with these subsistence practices were also found, such as iron arrowheads for hunting, iron sickles for harvesting, and wooden paddles for water-related activities. Further research is still needed to examine resource use across different time periods and to better understand how subsistence strategies responded to foreign influences and environmental changes.

@@

To explore changes in faunal remains over time, the original arbitrary levels of Kiwulan were assigned into six natural strata spanning from the 14th century to the 19th century, referred as cultural layers (CL1-Cl6) in this study. Each stratum was determined according to documentation of excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators. The diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks relating to the 19th century [@Wang2011; @Hsieh2009]. Although precisely identifying the stratum representing the specific time of European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above could still offer a relatively robust framework for comparing archaeological records before and after the foreign influence. Among these six temporal layers, the fourth cultural layer (CL4) can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second cultural layer (CL2) corresponds to the 19th century, associated with the large influx of Chinese migrants in Yilan.

@@

# Discussion
<!--results and chronology -->
The faunal assemblage from the major residential area at Kiwulan suggests a continuous reliance on sika deer from CL6 to CL2, followed by a shift to bovine in CL1. Despite deer were the main animal resource during all times, variations in quantity, ubiquity, butchering patterns, and proportions relative to other animals suggest changes relative to foreign presence. The cultural layers were grouped into four temporal sequences in relation to foreign presences for the following interpretation. The sequences include the pre-European period (CL6, CL5), the European period (CL4), the post-European period (CL3), and the Chinese period with large Han Chinese settled down (CL2, CL1). The therm "European period" does not imply direct colonial control over Kiwulan but rather the broad impact of European colonial activities in northern Taiwan, such as long-distance trade, tribute systems, and occasional attacks affecting Indigenous communities in Yilan. Although the small sample size across periods limits comprehensive interpretation, examining results in a continuous sequence still reveals distinct patterns in animal use.

<!--summary of overall observations across time periods -->
The sampled area at Kiwulan contains the large quantity of faunal remains from the European period, followed by the late pre-European period. Interestingly, deer bones from the late pre-European period presented the same ubiquity as in the European period but only half the quantity. This might indicate a consistent pattern of regular deer consumption at the site both before and after European presence, with intensified deer use during the European period. In addition to the dominance of deer, the increased relative proportion and ubiquity of suids in the European period suggests a more regular use of suids compared to other periods. The dramatic decline in deer bones after the European period may be associated with both a reduction in the Kiwulan population and the decrease of wild deer. Sika deer populations in western Taiwan largely declined from the 18th century due to habitat loss and overhunting [@Chen2000animal]. A slight increase in muntjac use suggests that they may have served as a substitute, replacing suids as a secondary animal resource from earlier periods. Despite these shifts, deer remained a primary resource for local consumption at Kiwulan during the post-European and early Chinese periods. Bovines were present in small numbers (yellow cattle) in the pre-European period, but became the sole large animal (water buffalo) used at Kiwulan in the late Chinese period. Since bovines were absent from prehistoric sites in Taiwan and are widely believed to have been introduced from mainland China in historical period [@Chen2000animal], their increasing presence suggests Chinese influence and aligns with the Indigenous relocation to eastern Taiwan since the 19th century due to Chinese population pressures [@Zhan1995].

<!--skeletal part frequencies and butchery patterns of deer-->
Differences in deer skeletal completeness across periods suggests changes in butchery locations and cooking practices. There were more complete carcasses in the European and late pre-European periods compared to the post-European and Chinese periods. Moreover, the substantial rise in fragmented head parts (cranium and mandible) during the European period indicates increased use of soft tissues, such as brains and tongues, or the transport of more intact deer to the settlement for processing. This is consistent with the historical account in 1603, Record of Formosa (Dongfan ji) [@Chen1603], which describes the consumption of deer tongues and group hunting practices that could facilitate the transportation of deer. Changes in cooking practices are also evident. The decline in burned bones over time, with the lowest percentage in the European period, suggests a shift from roasting to other cooking methods, such as boiling [@Lupo1997]. The size of chopped bones, which could fit into local ceramics at Kiwulan, might be associated with boiling practices. This could be further confirmed through organic residue analysis on pottery. Other meat preparation methods, such as drying through salting and smoking, are also possible [e.g., @Spyrou2019]. Although vertebrae and ribs, typically used for drying [@Friesen2001; @Binford1978], were present in low frequencies at Kiwulan, historical records mentioning the consumption or exchange of dried meat by the Kavalan people supports the possibility [@Chen1963]. The shift in cooking practices might result from the increased need for deer, leading people to adopt boiling for its benefits, such as maximizing the use of residual meat, facilitating grease rendering, and improving quality of lean meat [@Lupo1997], while drying helped preserve and transport the meat.

A consistent pattern of high frequencies of upper limb bones but low frequencies of vertebrae and ribs was identified across all periods. This might suggest selective butchering and a preference for meaty portions. Moreover, among the meaty portions, front quarters (humerus and radius) were consistently more abundant than hindquarters (femur and tibia). One possible explanation is that femurs and tibiae were heavily fragmented due to marrow extraction, as these bones have high marrow yields [@Madrigal2002; @Lupo1998], making them less identifiable in the assemblage. Evidence for marrow extraction is supported by at least 25% of long bones showing fractures and a moderate degree of fragmentation overall. Another explanation is that some hindquarters were used for trade, as Dutch records mention that deer legs and meat were commodities in addition to hides [@Blusse2006, pp.220-224], but they did not specify which legs. If deer legs, particularly hindquarters, were traded as bone-meat packages, this could account for the lower frequencies of hindquarter bones compared to front quarters. While this hypothesis requires further evidence, the overall high proportion of meaty portions at Kiwulan suggests that deer meat might be not only used for local consumption but also processed for trade. Thus, both factors could contribute to this pattern, rather than a single factor alone.

<!--deer hides-->
Deerskin trade had been prevalent in East Asia since the 14th century and was monopolized by Europeans in the 17th century when a large number of hides, especially from western Taiwan, were exported to Japan [@Cheng2017; @Hollmann1991; @Wang1981]. The cut marks identified on deer lower limbs at Kiwulan confirm the removal of deer skins, likely for trade with external traders. Moreover, the abundance of foreign goods at Kiwulan, including ornaments, porcelains, and metal items, further supports the frequent exchanges between local residents and outsiders [@Wang2020ornament; @Chen2007]. This suggests that northeastern Taiwan, like western Taiwan, was actively involved in and impacted by long-distance trade. Although lower limbs with skinning cuts did not show a substantial increase during the European period, the disproportionate rise in head parts might hint at the likelihood of brain tanning when preparing hides. Brain tanning is a traditional method for softening and preserving hides in many Indigenous cultures, including those of North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016]. In Taiwan, the present-day Bunun group in the central highlands has revived hide tanning based on oral history, primarily using peanuts or pig lard as tanning materials, with occasional use of animal brains [@Li1999, pp. 48; @Xu1994]. Despite the specific tanning materials used by the Kavalan people remain unknown due to the loss of traditions, brain tanning is a plausible method given the availability of deer heads from hunting.

<!-- population dynamics and arguments-->
The high population during the European period, as inferred by the frequencies of pottery, may have contributed to the overall increase in animal use [@Wang2020pot; @Hsieh2009]. However, the impact of foreign presence can be still be identified in deer butchery patterns, the relative proportions of animals, and animal ubiquity. First, lower limbs with skinning cuts were mostly found in the European period, followed by the late pre-European phase, aligning with the rise of regional trade in the 14th–15th centuries and its intensification in the 17th century. Second, if the high frequency of deer heads during the European period was solely due to food stress from population growth, leading to increased brain or tongue consumption, a parallel rise in marrow extraction would be expected. However, long bones with fracture patterns showing 10–30% less frequent during the European and late pre-European periods than in other periods. Third, population growth would have likely led to a more proportional increase in both large and small animals. Despite high taxonomic diversity during the European period, suids showed a disproportionate rise in ubiquity, quantity, and proportion. This indicates that the trade of deerskin and meat may have led Indigenous people to rely more on suids as an alternative prey due to their comparable body size and meat return rates to deer [@Broughton2011prey]. The role of suids as a secondary resource after deer has been observed in Indigenous cultures in western Taiwan [@Chuang2021], suggesting they may have served as a supplement when deer were scarce or limited, possibly because some deer were reserved for trade.
